---
title: "Race, genetics and IQ"
author: "Ian Brettell"
date: "5 January 2021"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    keep_md: true
---

# 1. Setup

* Working directory on EBI Cluster: `/hps/research1/birney/users/ian/rac_hyp`
* GitHub repository: <https://github.com/Ian-Brettell/racist_hypothesis>

## 1.1. Load R packages on local

```{r, message=FALSE}
library(here)
library(pegas)
library(tidyverse)
library(reshape2
library(readxl)
library(plotly
library(ggridges)
```

## 1.2. EBI cluster

### 1.2.1. Update GATK

```{r, engine='bash', eval = F}
cd /nfs/software/birney
wget https://github.com/broadinstitute/gatk/releases/download/4.1.4.1/gatk-4.1.4.1.zip
unzip gatk-4.1.4.1.zip

# amend aliases in ~/.bashrc and ~/.bash_profile
export PATH=$PATH:/nfs/software/birney/gatk-4.1.4.1/
```

## 1.3. Download 1GK data

### 1.3.1. Download from FTP

```{r, engine='bash', eval = F}
wget -r -p -k --no-parent -cut-dirs=5 ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/
```

### 1.3.2. Put list of files into list

```{bash, eval = F}
find vcfs/ftp.1000genomes.ebi.ac.uk/ALL.chr*.vcf.gz > racist_hypothesis/data/20200205_vcfs.list
```

### 1.3.3. Merge VCFs

```{bash, eval = F}
java -jar /nfs/software/birney/picard-2.9.0/picard.jar MergeVcfs \
  I=racist_hypothesis/data/20200205_vcfs.list \
  O=vcfs/1gk_all.vcf.gz
# Exception in thread "main" java.lang.IllegalArgumentException: The contig entries in input file /hps/research1/birney/users/ian/rac_hyp/vcfs/ftp.1000genomes.ebi.ac.uk/ALL.chrMT.phase3_callmom-v0_4.20130502.genotypes.vcf.gz are not compatible with the others.

# So remove that one from list above
sed -i '/MT/d' racist_hypothesis/data/20200205_vcfs.list

# run MergeVCFs again
java -jar /nfs/software/birney/picard-2.9.0/picard.jar MergeVcfs \
  I=racist_hypothesis/data/20200205_vcfs.list \
  O=vcfs/1gk_all.vcf.gz
  
# Exception in thread "main" java.lang.IllegalArgumentException: The contig entries in input file /hps/research1/birney/users/ian/rac_hyp/vcfs/ftp.1000genomes.ebi.ac.uk/ALL.chrY.phase3_integrated_v2a.20130502.genotypes.vcf.gz are not compatible with the others.
sed -i '/chrY/d' racist_hypothesis/data/20200205_vcfs.list

# run MergeVCFs again
java -jar /nfs/software/birney/picard-2.9.0/picard.jar MergeVcfs \
  I=racist_hypothesis/data/20200205_vcfs.list \
  O=vcfs/1gk_all.vcf.gz
# SUCCESS
```

## 1.4. Obtain GWAS data

### 1.4.1. Educational attainment

From **Lee et al. (2019)** Gene discovery and polygenic prediction from a genome-wide association study of educational attainment in 1.1 million individuals, *Nature*: <https://www.nature.com/articles/s41588-018-0147-3>.

Data download link: <https://static-content.springer.com/esm/art%3A10.1038%2Fs41588-018-0147-3/MediaObjects/41588_2018_147_MOESM3_ESM.xlsx>

Saved here: `data/20180723_Lee-et-al_supp-tables.xlsx`

### 1.4.2. Height

From **Yengo et al. (2018)** Meta-analysis of genome-wide association studies for height and body mass index in approximately 700000 individuals of European ancestry, *Human Molecular Genetics*: <https://academic.oup.com/hmg/article-abstract/27/20/3641/5067845>.

Data downloaded from this webpage:
<https://portals.broadinstitute.org/collaboration/giant/index.php/GIANT_consortium_data_files>
Data download link: <https://portals.broadinstitute.org/collaboration/giant/images/e/e2/Meta-analysis_Locke_et_al%2BUKBiobank_2018_top_941_from_COJO_analysis_UPDATED.txt.gz> 

```{bash, eval = F}
cd racist_hypothesis/data
# download
wget https://portals.broadinstitute.org/collaboration/giant/images/e/e2/Meta-analysis_Locke_et_al%2BUKBiobank_2018_top_941_from_COJO_analysis_UPDATED.txt.gz
# unzip
gunzip Meta-analysis_Locke_et_al%2BUKBiobank_2018_top_941_from_COJO_analysis_UPDATED.txt.gz
# rename
mv Meta-analysis_Wood_et_al+UKBiobank_2018_top_3290_from_COJO_analysis.txt 20181015_Yengo-et-al_snps_height.txt
```

Saved here: `data/20181015_Yengo-et-al_snps_height.txt`

### 1.4.3. IBD

From **Huang et al. (2017)** Fine-mapping inflammatory bowel disease loci to single-variant resolution, *Nature*: <https://www.nature.com/articles/nature22969#Sec29>.

Data download link (Supplementary Table 1): <https://static-content.springer.com/esm/art%3A10.1038%2Fnature22969/MediaObjects/41586_2017_BFnature22969_MOESM2_ESM.xlsx>

Saved here: `data/20170628_Huang-et-al_supp-table-1.xlsx`

### 1.4.4. Skin pigmentation

* **Crawford et al. (2017)** Loci associated with skin pigmentation identified in African populations, *Science*: <https://science.sciencemag.org/content/358/6365/eaan8433.abstract?casa_token=6tGEjct1nUQAAAAA:IL7LCz-xQ9l6rLhxBk5VcGBjwTrEa5UpAlC-nCl2mvcASu4iZbWMiu_Uj8YUHIISgCibOd1ya25sOQ>. Take table 1, and manually transform 'Ancestral>Divided' column into new columns titled 'tested_allele' (the allele in bold) and 'other_allele'. Save here: `data/20171117_Crawford-et-al_Table-1.xlsx`
* **Adhikari et al. (2019)** A GWAS in Latin Americans highlights the convergent evolution of lighter skin pigmentation in Eurasia, *Nature*: <https://www.nature.com/articles/s41467-018-08147-0>. "Summary statistics from the GWAS analyses is deposited at GWAS central with the link <http://www.gwascentral.org/study/HGVST3308>". Under the 'Association results' tab, there is one dataset for each of the 6 phenotypes tested:  
  - Melanin index
  - Hair color
  - Eye color
  - Digital eye color: L (lightness)
  - Digital eye color: C (chroma)
  - Digital eye color: cosH (cosine of hue)
Difficult to ascertain which ones had genome-wide significance.
Instead, pull tables directly from paper and supplementary materials, and put here in different sheets: `data/20190121_Adhikari-et-al_snps.xlsx`
  - *Table 1*: 18 lead SNPs from paper, each with a different p-value for one of the 6 phenotypes.
  - *supp_table_6*: 11 SNPs associated with combined traits.
  - *supp_table_12*: 161 SNPs collagted from published association studies on pigmentation. See table for references, which were used to identify other pigmentation GWAS studies.
* **Hernandez-Pacheco et al. (2017)** Identification of a novel locus associated with skin colour in African-admixed populations, *Scientific Reports*: <https://www.nature.com/articles/srep44548>. 9 hits with genome-wide significance here: `data/20170316_Hernandez-Pacheco-et-al.xlsx`.

Others compiled into the single XLSX doc `data/20200622_pigmentation_snps.xlsx`:

* 20190321_Jonnalagadda-et-al: **Jonnalagadda et al. (2019)** A Genome-Wide Association Study of Skin and Iris Pigmentation among Individuals of South Asian Ancestry, *Genome Biology and Evolution*: <https://academic.oup.com/gbe/article/11/4/1066/5416147>. Took 9 SNPs associated with iris colour from Table 1, and 14 SNPs described in previous studies from Table 2.
* 20171130_Martin-et-al: **Martin et al. (2017)** An Unexpectedly Complex Architecture for Skin Pigmentation in Africans, *Cell*: <https://www.cell.com/cell/fulltext/S0092-8674%2817%2931324-7>. Took all 42 SNPs from Table S6A in Supplemental Information here: <https://www.cell.com/cms/10.1016/j.cell.2017.11.015/attachment/eccead83-1a96-4444-9032-f968ee481d15/mmc2.xlsx>.
* 20150512_Liu-et-al: **Liu et al. (2015)** Genetics of skin color variation in Europeans: genome-wide association studies with functional follow-up, *Human Genetics*: <https://link.springer.com/article/10.1007%2Fs00439-015-1559-0>. Took all 9 SNPs from Table 1.
* 20121031_Candille-et-al: **Candille et al. (2012)** Genome-wide association studies of quantitatively measured skin, hair, and eye pigmentation in four European populations, *PLoS One*: <https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0048294>. Took all 8 SNPs from Table 2.
* 20100614_Gerstenblith-et-al: **Gerstenblith et al. (2010)** Genome-wide association studies of pigmentation and skin cancer: a review and meta-analysis, *Pigment Cell Melanoma Research*: <https://onlinelibrary.wiley.com/doi/full/10.1111/j.1755-148X.2010.00730.x>. Took all 39 SNPs from Table 4.
* 20080516_Han-et-al: **Han et al. (2008)** A Genome-Wide Association Study Identifies Novel Alleles Associated with Hair Color and Skin Pigmentation, *PLOS Genetics*: <https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000074>. Took all 38 SNPs from Table 2. 
  
### 1.4.5. Create list of target SNPs
  
```{r}
pig_snps <- list()
# Crawford
pig_snps[["crawford"]] <- readxl::read_excel(here("data", "20171117_Crawford-et-al_Table-1.xlsx")) %>% 
  dplyr::select(rsid = "RSID", everything()) %>% 
  dplyr::filter(!is.na(rsid),
                !rsid == ".")

# Adhikari
pig_snps[["adhikari_tbl_1"]] <- readxl::read_excel(here("data", "20190121_Adhikari-et-al_snps.xlsx"),
                                                   sheet = "Table 1",
                                                   skip = 1) %>% 
  dplyr::select(rsid = "rsID", everything()) %>% 
  dplyr::filter(!is.na(rsid))

pig_snps[["adhikari_supp_6"]] <- readxl::read_excel(here("data", "20190121_Adhikari-et-al_snps.xlsx"),
                                                   sheet = "supp_table_6") %>% 
  dplyr::select(rsid = "SNP", everything()) %>% 
  dplyr::filter(!is.na(rsid))  

pig_snps[["adhikari_supp_12"]] <- readxl::read_excel(here("data", "20190121_Adhikari-et-al_snps.xlsx"),
                                                   sheet = "supp_table_12") %>% 
  dplyr::select(rsid = "SNP", everything()) %>% 
  dplyr::filter(!is.na(rsid))

# Hernandez-Pacheco
pig_snps[["hernandez-pacheco"]] <- readxl::read_excel(here("data", "20170316_Hernandez-Pacheco-et-al.xlsx"))

# Doc with SNPs from multiple studies
sheet_names <- readxl::excel_sheets(here("data", "20200622_pigmentation_snps.xlsx"))
compiled_snps <- lapply(sheet_names, function(x){
  x <- readxl::read_excel(here("data", "20200622_pigmentation_snps.xlsx"),
                     sheet = x)
})
names(compiled_snps) <- sheet_names

# Combine
pig_snps <- c(pig_snps, compiled_snps)
```

#### 1.4.5.1. Pull out unique SNPs

```{r}
pig_df <- lapply(pig_snps, function(x) dplyr::select(x, rsid))
pig_df <- dplyr::bind_rows(pig_df)
pig_df <- unique(pig_df)

nrow(pig_df)
head(pig_df)
```

## 1.5. Filter 1GK VCF for target SNPs

### 1.5.1. EA

#### 1.5.1.1. Create list of target SNPs

```{r, eval = F}
# extract from excel doc
snps_eduyrs <- read_xlsx(here::here("data", "20180723_Lee-et-al_supp-tables.xlsx"), sheet = "2. EduYears Lead SNPs", skip = 1, n_max = 1271)
# write table of SNPs
write.table(snps_eduyrs[["SNP"]], here::here("data", "20200316_snps_eduyears.list"), quote = F, row.names = F, col.names = F)
```

#### 1.5.1.2. Create filtered VCF

```{bash, eval = F}
gatk SelectVariants \
  -R refs/hs37d5.fa.gz \
  -V vcfs/1gk_all.vcf.gz \
  --keep-ids racist_hypothesis/data/20200316_snps_eduyears.list \
  -O vcfs/snphits_eduyrs.vcf.gz
```

### 1.5.2. Height

#### 1.5.2.1. Create list of target SNPs

```{bash, eval = F}
cut -f1 racist_hypothesis/data/20181015_Yengo-et-al_snps_height.txt | tail -n+2 > racist_hypothesis/data/20200318_snps_height.list
```

#### 1.5.2.2. Extract calls for those SNPs from VCF

```{bash, eval = F}
gatk SelectVariants \
  -R refs/hs37d5.fa.gz \
  -V vcfs/1gk_all.vcf.gz \
  --keep-ids racist_hypothesis/data/20200318_snps_height.list \
  -O vcfs/snphits_height.vcf.gz
```

### 1.5.3. IBD

#### 1.5.3.1. Create list of target SNPs

```{r, eval = F}
# extract from excel doc
snps_ibd <- read_xlsx(path = here::here("data", "20170628_Huang-et-al_supp-table-1.xlsx"),
                      sheet = "list of variants", )
# write tables of SNPs
write.table(x = snps_ibd$variant,
            file = here::here("data", "20200319_snps_ibd.list"),
            quote = F,
            row.names = F,
            col.names = F)
```

#### 1.5.3.2. Extract calls for those SNPs from VCF

```{bash, eval = F}
gatk SelectVariants \
  -R refs/hs37d5.fa.gz \
  -V vcfs/1gk_all.vcf.gz \
  --keep-ids racist_hypothesis/data/20200319_snps_ibd.list \
  -O vcfs/snphits_ibd_full.vcf.gz
```

### 1.5.4. Skin pigmentation

#### 1.5.4.1. Create list of target SNPs

```{r, eval = F}
write.table(x = pig_df$rsid,
            file = here::here("data", "20200622_snps_pig.list"),
            quote = F,
            row.names = F,
            col.names = F)
```

#### 1.5.4.2. Extract calls for those SNPs from VCF

```{bash, eval = F}
gatk SelectVariants \
  -R refs/hs37d5.fa.gz \
  -V vcfs/1gk_all.vcf.gz \
  --keep-ids racist_hypothesis/data/20200622_snps_pig.list \
  -O vcfs/snphits_pig.vcf.gz
```

## 1.6. Get allele frequencies with `Plink2`

### 1.6.1. Import 1GK metadata (for sample-population key)

Downloded via this page: <http://www.internationalgenome.org/data>
Download link: <http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/technical/working/20130606_sample_info/20130606_sample_info.xlsx>. 

```{r, warning=FALSE}
meta <- read_xlsx(here::here("data", "20130606_sample_info.xlsx"), sheet = "Sample Info") %>% dplyr::select(Sample, Population, Gender)
meta
```

### 1.6.2. Write population file for Plink2

```{r, eval = F}
write.table(meta[, 1:2],
            here::here("data", "plink2_sample_popn_key.txt"),
            quote = F,
            sep = "\t",
            row.names = F,
            col.names = F)
```

### 1.6.3. Set up directories on EBI cluster

```{bash, eval = F}
mkdir racist_hypothesis/data/20200622_plink2_alfreqs
mkdir racist_hypothesis/data/20200622_plink2_alfreqs/hei
mkdir racist_hypothesis/data/20200622_plink2_alfreqs/edu
mkdir racist_hypothesis/data/20200622_plink2_alfreqs/ibd
mkdir racist_hypothesis/data/20200622_plink2_alfreqs/pig
```

### 1.6.4. Run Plink2

```{bash, eval = F}
# Edu Years
/nfs/software/birney/plink2.3/plink2 \
  --vcf vcfs/snphits_eduyrs.vcf.gz \
  --freq \
  --max-alleles 2 \
  --pheno iid-only racist_hypothesis/data/plink2_sample_popn_key.txt \
  --loop-cats PHENO1 \
  --out racist_hypothesis/data/20200622_plink2_alfreqs/edu/edu

# Height
/nfs/software/birney/plink2.3/plink2 \
  --vcf vcfs/snphits_height.vcf.gz \
  --freq \
  --max-alleles 2 \
  --pheno iid-only racist_hypothesis/data/plink2_sample_popn_key.txt \
  --loop-cats PHENO1 \
  --out racist_hypothesis/data/20200622_plink2_alfreqs/hei/hei
  
# IBD
/nfs/software/birney/plink2.3/plink2 \
  --vcf vcfs/snphits_ibd.vcf.gz \
  --freq \
  --max-alleles 2 \
  --pheno iid-only racist_hypothesis/data/plink2_sample_popn_key.txt \
  --loop-cats PHENO1 \
  --out racist_hypothesis/data/20200622_plink2_alfreqs/ibd/ibd

# Pigmentation
/nfs/software/birney/plink2.3/plink2 \
  --vcf vcfs/snphits_pig.vcf.gz \
  --freq \
  --max-alleles 2 \
  --pheno iid-only racist_hypothesis/data/plink2_sample_popn_key.txt \
  --loop-cats PHENO1 \
  --out racist_hypothesis/data/20200622_plink2_alfreqs/pig/pig
```

# 2. Analysis

## 2.1. Compare allele frequencies

### 2.1.1. Read in data

```{r}
target_dirs <- list.dirs(here::here("data", "20200622_plink2_alfreqs"), recursive = F)

al_freq_lst <- lapply(target_dirs, function(x){
  target_files <- list.files(x, pattern = ".afreq", full.names = T)
  # read in data
  data_lst <- lapply(target_files, function(target_file){
    read.table(target_file,
               header = T,
               comment.char = "")
  })
  # fix names of populations
  names(data_lst) <- gsub(pattern = "edu.|hei.|pig.|.afreq",
                          replacement = "",
                          x = list.files(x, pattern = ".afreq"))
  return(data_lst)
})

# set names
names(al_freq_lst) <- basename(target_dirs)
```

### 2.1.2. Turn into single table for each pheno

```{r}
al_freq_df <- lapply(al_freq_lst, function(pheno){
  out <- dplyr::bind_rows(pheno, .id = "population") %>% 
    tidyr::pivot_wider(id_cols = c(X.CHROM, ID, REF, ALT),
                       names_from = population,
                       values_from = ALT_FREQS)
})
```

### 2.1.3. Randomly swap minor allele

```{r}
set.seed(65)
rdm_sds <- sample(1:100, 3)

counter <- 0
al_freq_df_shuff <- lapply(al_freq_df, function(pheno){
  counter <<- counter + 1
  # set seed
  set.seed(rdm_sds[counter])
  # select SNPs to swap (half of total)
  tgt_indcs <- sample(nrow(pheno), nrow(pheno) /2)
  # swap minor alleles
  pheno[tgt_indcs, 5:ncol(pheno)] <- 1 - pheno[tgt_indcs, 5:ncol(pheno)]
  # return pheno
  return(pheno)
})
```

### 2.1.4. Plot

```{r}
# Set up titles vector
titles <- c("Educational Attainment", "Height", "Pigmentation")
```

#### 2.1.4.1. D

##### 2.1.4.1.1. YRI v CHS

```{r, message=F, warning=F}
counter <- 0
lapply(al_freq_df_shuff, function(pheno){
  counter <<- counter + 1
  ggplot(pheno,
         aes(YRI, CHS)) +
    geom_point(size = 0.5) +
    coord_fixed() +
    geom_smooth(se = F, colour = "red") +
    geom_abline(intercept = 0, slope = 1, colour = "blue") +
    xlab("Allele frequency in YRI") +
    ylab("Allele frequency in CHS") +
    labs(title = titles[counter])
})
```

##### 2.1.4.1.2. YRI v CEU

```{r, message=F, warning=F}
counter <- 0
lapply(al_freq_df_shuff, function(pheno){
  counter <<- counter + 1
  ggplot(pheno,
         aes(YRI, CEU)) +
    geom_point(size = 0.5) +
    coord_fixed() +
    geom_smooth(se = F, colour = "red") +
    geom_abline(intercept = 0, slope = 1, colour = "blue") +
    xlab("Allele frequency in YRI") +
    ylab("Allele frequency in CEU") +
    labs(title = titles[counter])
})
```

#### 2.1.4.2. D

```{r, message=F, warning=F}
colourscales <- c("Viridis", "Hot", "Electric")
titles <- c("Educational Attainment", "Height", "Skin/hair pigmentation")

counter <- 0
plts <- lapply(al_freq_df_shuff, function(pheno){
  counter <<- counter + 1
  # set graph resolution
  graph_reso <- 0.05
  # get lm for data
  loess_model <- loess(CEU ~ 0 + CHS + YRI, data = pheno)
  # set up axes
  axis_x <- seq(min(pheno$CHS), max(pheno$CHS), by = graph_reso)
  axis_y <- seq(min(pheno$YRI), max(pheno$YRI), by = graph_reso)
  # sample points
  lm_surface <- expand.grid(CHS = axis_x,
                            YRI = axis_y,
                            KEEP.OUT.ATTRS = F)
  lm_surface$CEU <- predict(loess_model, newdata = lm_surface)
  lm_surface <- acast(lm_surface, YRI ~ CHS, value.var = "CEU")
  # create plot
  plt <- plot_ly(pheno,
                 x = ~CHS,
                 y = ~YRI,
                 z = ~CEU,
                 type = "scatter3d",
                 mode = "markers",
                 marker = list(size = 2),
                 text = pheno$ID) 
  plt <- add_trace(plt,
                   z = lm_surface,
              x = axis_x,
              y = axis_y,
              type = "surface",
              colorscale = colourscales[counter]) %>% 
    layout(title = titles[counter])
  return(plt)
})

plts$edu
plts$hei
plts$pig
```

## 2.2. Fst

### 2.2.1. Find target VCFs

```{r}
# list target VCFs
target_vcfs <- list.files(here::here("data"),
                          pattern = glob2rx("snphits_*.gz"), 
                          full.names = T)

# filter for the three we want
target_vcfs <- target_vcfs[grep("eduyrs|height|pig", target_vcfs)]


```

### 2.2.2. With all populations

#### 2.2.2.1. Get Fst stats

```{r, message=F, warning=F, results=F}
# Create raw list of variants
vcf_list_raw <- lapply(target_vcfs, function(vcf_file){
  vcf_out <- pegas::read.vcf(vcf_file)
})

# Create vector of populations
populations <- unlist(lapply(rownames(vcf_list_raw[[1]]), function(sample){
  meta$Population[meta$Sample == sample]
}))

# Generate Fst stats
fst_out_lst <- lapply(vcf_list_raw, function(pheno){
  as.data.frame(pegas::Fst(pheno, pop = populations))
})

# make rownames into separate column
fst_out_lst <- lapply(fst_out_lst, function(pheno){
  pheno$snp <- rownames(pheno)
  return(pheno)
})
names(fst_out_lst) <- titles

# bind into single DF
fst_out_df <- dplyr::bind_rows(fst_out_lst, .id = "phenotype")
head(fst_out_df)
```

#### 2.2.2.2. Plot density

##### 2.2.2.2.1. D

```{r, warning= F}
ggplot(fst_out_df, aes(Fst, fill = phenotype)) +
  geom_density(alpha = 0.7) +
  labs(fill = "Phenotype") +
  ylab("Density") +
  theme_bw() +
  scale_fill_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"))
```

##### 2.2.2.2.2. D

```{r, warning = F}
# factorise 
fst_out_df$phenotype <- factor(fst_out_df$phenotype,
                                    levels = c("Skin/hair pigmentation", "Height", "Educational Attainment"))

ggplot() +
  geom_density_ridges2(data = fst_out_df,
                       mapping = aes(x = Fst, y = phenotype, fill = phenotype),
                       scale = 2) +
  scale_fill_manual(values = c("#FC4E07", "#00AFBB", "#E7B800")) +
  ylab(label = NULL) +
  theme_bw() +
  guides(fill = guide_legend(reverse=T, 
                             title = "Phenotype")) +
  scale_y_discrete(expand = expand_scale(add = c(0.2, 2.3)))
```

### 2.2.3. With just YRI, CEU, and CHS

#### 2.2.3.1. Get Fst stats

```{r}
# get samples from target popns only
target_popns <- which(populations %in% c("YRI", "CEU", "CHS"))
populations_3pop <- populations[target_popns]

vcf_list_raw_3pop <- lapply(vcf_list_raw, function(pheno){
  pheno[target_popns, ]
})

# Generate Fst stats
fst_out_lst_3pop <- lapply(vcf_list_raw_3pop, function(pheno){
  as.data.frame(pegas::Fst(pheno, pop = populations_3pop))
})

# make rownames into separate column
fst_out_lst_3pop <- lapply(fst_out_lst_3pop, function(pheno){
  pheno$snp <- rownames(pheno)
  return(pheno)
})
names(fst_out_lst_3pop) <- titles

# bind into single DF
fst_out_df_3pop <- dplyr::bind_rows(fst_out_lst_3pop, .id = "phenotype")
head(fst_out_df_3pop)
```

#### 2.2.3.2. Plot density

##### 2.2.3.2.1. D

```{r, warning=F}
ggplot(fst_out_df_3pop, aes(Fst, fill = phenotype)) +
  geom_density(alpha = 0.7) +
  labs(fill = "Phenotype") +
  ylab("Density") +
  theme_bw() +
  scale_fill_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"))
```

##### 2.2.3.2.2. D

```{r, warning=F}
# factorise 
fst_out_df_3pop$phenotype <- factor(fst_out_df_3pop$phenotype,
                                    levels = c("Skin/hair pigmentation", "Height", "Educational Attainment"))

ggplot() +
  geom_density_ridges2(data = fst_out_df_3pop,
                       mapping = aes(x = Fst, y = phenotype, fill = phenotype),
                       scale = 2) +
  scale_fill_manual(values = c("#FC4E07", "#00AFBB", "#E7B800")) +
  ylab(label = NULL) +
  theme_bw() +
  guides(fill = guide_legend(reverse=T, 
                             title = "Phenotype")) +
  scale_y_discrete(expand =p expand_scale(add = c(0.2, 2.3)))
```