---
title: "$F_{ST}$ variation across human traits"
author: "Ian Brettell"
date: '`r format(Sys.Date())`'
output:
  html_document:
    toc: true
    toc_float: true
    dev: 'svg'
    number_sections: true
    keep_md: false
    pandoc_args: --lua-filter=color-text.lua
    highlight: pygments
#output: html_notebook
#editor_options: 
#  chunk_output_type: inline
---


* [Working directory]{color="#4f0943"} on EBI Codon HPC: `/hps/nobackup/birney/users/ian/hmn_fst`
* [GitHub repository]{color="#4f0943"}: <https://github.com/brettellebi/human_traits_fst>


# Source libraries, functions and plotting parameters

```{r, warning = F, message = F}
library(here)

source(here::here("code", "scripts", "source.R"))
```

# Gather and process 1KG VCFs and GWAS Catalog data

Full Snakemake pipeline here: <https://github.com/brettellebi/human_traits_fst/tree/master/code/snakemake/20210416>

```{bash, eval = F}
cd /hps/software/users/birney/ian/repos/human_traits_fst
conda activate snakemake
sing_load
snmk_proj="20210416"

snakemake \
  --jobs 5000 \
  --latency-wait 500 \
  --cluster-config code/snakemake/$snmk_proj/config/cluster.json \
  --cluster 'bsub -g /snakemake_bgenie -J {cluster.name} -n {cluster.n} -M {cluster.memory} -o {cluster.output} -e {cluster.error}' \
  --keep-going \
  --rerun-incomplete \
  --use-conda \
  --use-singularity \
  -s code/snakemake/$snmk_proj/Snakefile \
  -p
```

# Read in data and post-process

```{r, message = F, results = "hide"}
# Variables
gcat_traits = gwasrapidd::get_traits(efo_trait = target_traits)
trait_ids = gcat_traits@traits$efo_id
names(trait_ids) = trait_ids
target_dir = here::here("data/gwasrapidd/20210525")

# Read in raw associations objects
assocs_raw = lapply(trait_ids, function(TRAIT_ID){
  readRDS(file.path(target_dir, "associations_raw", paste(TRAIT_ID, ".rds", sep = "")))
})

# Read in raw studies objects
studies_raw = lapply(trait_ids, function(TRAIT_ID){
  readRDS(file.path(target_dir, "studies_raw", paste(TRAIT_ID, ".rds", sep = "")))
})

# Read in association-study key
studies_key = lapply(trait_ids, function(TRAIT_ID){
  readRDS(file.path(target_dir, "studies_key", paste(TRAIT_ID, ".rds", sep = "")))  
})

# Read in genotypes from VCFs
genos_raw = lapply(trait_ids, function(TRAIT_ID){
  path = file.path(target_dir, "vcfs", paste(TRAIT_ID, ".vcf.gz", sep = ""))
  n_variants = nrow(pegas::VCFloci(path))
  pegas::read.vcf(path, to = n_variants)
})

# Read in loci from VCFs
loci_info = lapply(trait_ids, function(TRAIT_ID){
  out = pegas::VCFloci(file.path(target_dir, "vcfs", paste(TRAIT_ID, ".vcf.gz", sep = "")))
  out %>%
    dplyr::mutate(CHROM = as.double(CHROM))
})

# Create vector of populations
## Read in population file
pop_file = readr::read_csv(here::here("data/1kg_populations.csv"))
## Create vector of populations
populations = unlist(lapply(rownames(genos_raw[[1]]), function(sample){
    pop_file$Population[pop_file$Sample == sample]
}))
## Create data frame with just 2,504 individuals
pops_1kg = pop_file %>% 
  dplyr::filter(Sample %in% rownames(genos_raw[[1]]))

# Read in clumped SNPs
clumped = lapply(trait_ids, function(TRAIT_ID){
  readr::read_delim(file.path(target_dir, "plink/clumped", paste(TRAIT_ID, ".clumped", sep = "")),
                    delim = " ", trim_ws = T)
})

# Read in 012 genotypes
genos_012 = lapply(trait_ids, function(TRAIT_ID){
  readr::read_tsv(file.path(target_dir, "plink/recode_012", paste(TRAIT_ID, ".traw", sep = ""))) %>% 
    # rename columns to remove duplicated ID
    dplyr::rename_with(~str_split(.x, pattern = "_", simplify = T) %>% 
                       subset(select = 1),
                       .cols = matches("_"))
})

```

## Counts

How many unique SNPs for each trait?

```{r}
# How many unique SNPs in associations?
raw_assocs_count = purrr::map(assocs_raw, function(TRAIT_ID){
  TRAIT_ID@risk_alleles %>% 
    dplyr::select(variant_id)
}) %>% 
  dplyr::bind_rows(.id = "efo_id") %>% 
  dplyr::left_join(.,
                   gcat_traits@traits,
                   by = "efo_id") %>% 
  # combine various pigmentation traits
  dplyr::mutate(TRAIT = dplyr::recode(trait, !!!pig_recode_vec),
                TRAIT = factor(TRAIT, levels = extended_traits)) %>% 
  dplyr::group_by(TRAIT) %>% 
  dplyr::distinct(variant_id) %>% 
  dplyr::count(name = "N_ASSOCS")

#How many unique SNPs in VCFs
purrr::map(loci_info, function(TRAIT_ID){
  TRAIT_ID %>% 
    dplyr::select(ID)
}) %>% 
  dplyr::bind_rows(.id = "efo_id") %>% 
  dplyr::left_join(.,
                   gcat_traits@traits,
                   by = "efo_id") %>% 
  dplyr::mutate(TRAIT = dplyr::recode(trait, !!!pig_recode_vec),
                TRAIT = factor(TRAIT, levels = extended_traits)) %>% 
  dplyr::group_by(TRAIT) %>% 
  dplyr::distinct(ID) %>% 
  dplyr::count(name = "N_VCF") %>% 
  # bind with raw assocs counts
  dplyr::left_join(raw_assocs_count,
                   .,
                   by = "TRAIT") %>% 
  # Order and add column with percentage
  dplyr::mutate(PERC_RETREIVED = N_VCF / N_ASSOCS) %>% 
  dplyr::arrange(.by_group = T) %>% 
  knitr::kable(.)


```

## Calculate Fst with `pegas`

```{r, eval = F}
fst_pegas = lapply(genos_raw, function(TRAIT_ID){
  pegas::Fst(TRAIT_ID, pop = populations)
})
```

```{r, eval = F, echo = F}
out_path = here::here("data/gwasrapidd/20210525/pegas/fst/fst.rds")
dir.create(dirname(out_path), showWarnings = F, recursive = T)
saveRDS(fst_pegas, out_path)
```

```{r, echo = F}
fst_pegas = readRDS(here::here("data/gwasrapidd/20210525/pegas/fst/fst.rds"))
```

## Bind relevant data

```{r}
fst_pegas_df = lapply(trait_ids, function(TRAIT_ID){
  # Loci info
  loci = loci_info[[TRAIT_ID]] %>% 
    dplyr::select(CHROM, POS, ID, REF, ALT)
  # Risk alleles
  risk_alleles = assocs_raw[[TRAIT_ID]]@risk_alleles %>% 
    dplyr::select(association_id, variant_id, risk_allele)
  # Study key
  study_key = studies_key[[TRAIT_ID]]
  # Studies
  studies = studies_raw[[TRAIT_ID]]@studies %>% 
    dplyr::select(study_id, initial_sample_size, replication_sample_size)
  # Fst
  fst = fst_pegas[[TRAIT_ID]] %>%
    data.frame(.) %>% 
    tibble::rownames_to_column(var = "ID") %>% 
    dplyr::select(ID, FST_PEGAS = Fst)
  
  # Bind all
  out = dplyr::left_join(loci, risk_alleles, by = c("ID" = "variant_id")) %>% 
    dplyr::left_join(study_key, by = "association_id") %>% 
    dplyr::left_join(studies, by = "study_id") %>% 
    dplyr::left_join(fst, by = "ID")
}) %>% 
  dplyr::bind_rows(.id = "EFO_ID") %>%
  # Add `trait` for each `efo_id` from `gcat_traits` key
  dplyr::left_join(gcat_traits@traits %>% 
                     dplyr::select(efo_id, trait),
                   by = c("EFO_ID" = "efo_id")) %>% 
  dplyr::select(EFO_ID, trait, everything()) %>% 
  # Recode Pigmentation traits
  dplyr::mutate(TRAIT = dplyr::recode(trait, !!!pig_recode_vec)) %>% 
  # Factorise to order
  dplyr::mutate(TRAIT = factor(TRAIT, levels = extended_traits))

# Create a second df with clumped SNPs
## Get clumped SNPs
clumped_snps = lapply(clumped, function(TRAIT_ID) {
  TRAIT_ID %>%
    dplyr::pull(SNP)
}) %>% purrr::flatten_chr()
## Create list and bind
fst_pegas_df_clumps= list(
  "ALL" = fst_pegas_df,
  "CLUMPED" = fst_pegas_df %>% 
    dplyr::filter(ID %in% clumped_snps)
) %>% 
  dplyr::bind_rows(.id = "FILTER")

# How many unique SNPs in total (having grouped by TRAIT)
fst_pegas_df_clumps %>% 
  dplyr::filter(FILTER == "ALL") %>% 
  dplyr::group_by(TRAIT) %>% 
  dplyr::distinct(ID) %>% 
  nrow()

# How many unique SNPs in total after clumping (having grouped by TRAIT)
fst_pegas_df_clumps %>% 
  dplyr::filter(FILTER == "CLUMPED") %>% 
  dplyr::group_by(TRAIT) %>% 
  dplyr::distinct(ID) %>% 
  nrow()

```

# Plot

## Initial counts of variants

```{r}
fst_pegas_df_clumps %>%
  # create new column combining skin pigmentation traits
  dplyr::mutate(trait_with_pig = dplyr::recode(trait,
                                               "skin pigmentation" = "skin pigmentation / measurement",
                                               "skin pigmentation measurement" = "skin pigmentation / measurement")) %>% 
  # order traits
  dplyr::mutate(trait_with_pig = factor(trait_with_pig, levels = traits_with_pig)) %>%
  # remove all pigmentation traits other than skin pigmentation and skin pigmentation measurement
  dplyr::filter(trait_with_pig %in% traits_with_pig) %>% 
  # group by FILTER and TRAIT to take unique SNPs within each group
  dplyr::group_by(FILTER, trait_with_pig) %>% 
  # take unique IDs
  dplyr::distinct(ID, .keep_all = T) %>% 
  # ungroup
  dplyr::ungroup() %>% 
  # plot
  ggplot() +
    geom_bar(aes(trait_with_pig, fill = trait_with_pig)) +
    scale_fill_manual(values = pal_primary_with_pig) +
    theme_bw() +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank()) +
    ylab("N loci") +
    facet_wrap(~FILTER, nrow = 2)
```

## $F_{ST}$ ridges

```{r, fig.cap = "$F_{ST}$ calculated with 26 populations", fig.height=10, fig.width=8}
fst_pegas_df_clumps %>% 
  # filter for clumped index SNPs
  dplyr::filter(FILTER == "CLUMPED") %>%
  # create new column combining skin pigmentation traits
  dplyr::mutate(trait_with_pig = dplyr::recode(trait,
                                               "skin pigmentation" = "skin pigmentation / measurement",
                                               "skin pigmentation measurement" = "skin pigmentation / measurement")) %>% 
  # remove all pigmentation traits other than skin pigmentation and skin pigmentation measurement
  dplyr::filter(trait_with_pig %in% traits_with_pig) %>% 
  # group by trait to take unique SNPs
  dplyr::group_by(trait_with_pig) %>% 
  # take unique IDs
  dplyr::distinct(ID, .keep_all = T) %>%
  # reverse order of traits to put `body height` at the top
  dplyr::mutate(trait_with_pig = factor(trait_with_pig, levels = rev(traits_with_pig))) %>%
  # plot
  ggplot(aes(FST_PEGAS, trait_with_pig, fill = trait_with_pig, colour = trait_with_pig)) +
    geom_density_ridges(scale = 2,
                        bandwidth = 0.003,
                        calc_ecdf = TRUE,
                        quantiles = c(0.5, 0.9),
                        quantile_lines = T,
                        jittered_points = T,
                        point_shape = '|', alpha = 0.85, point_size = 2,
                        position = position_points_jitter(height = 0),
    ) +
  scale_fill_manual(values = pal_primary_with_pig) +
  scale_colour_manual(values = karyoploteR::darker(pal_primary_with_pig)) +
  guides(fill = F, colour = F) +
  theme_bw() +
  scale_y_discrete(expand = expansion(add = c(0.2, 2.3))) +
  ylab(NULL) +
  ggtitle("Median and 90%") 
```

## $F_{ST}$ eCDF 

```{r, fig.height=8, fig.width=8, fig.cap = "Faceted"}
fst_pegas_df_clumps %>% 
  # filter for clumped index SNPs
  dplyr::filter(FILTER == "CLUMPED") %>%

  # create new column combining skin pigmentation traits
  dplyr::mutate(trait_with_pig = dplyr::recode(trait,
                                               "skin pigmentation" = "skin pigmentation / measurement",
                                               "skin pigmentation measurement" = "skin pigmentation / measurement")) %>% 
  # remove all pigmentation traits other than skin pigmentation and skin pigmentation measurement
  dplyr::filter(trait_with_pig %in% traits_with_pig) %>% 
  # group by trait to take unique SNPs
  dplyr::group_by(trait_with_pig) %>% 
  # take unique IDs
  dplyr::distinct(ID, .keep_all = T) %>%
  # order traits
  dplyr::mutate(trait_with_pig = factor(trait_with_pig, levels = traits_with_pig)) %>%
  # plot
  ggplot() +
    stat_ecdf(aes(FST_PEGAS, colour = trait_with_pig)) +
    scale_colour_manual(values = pal_primary_with_pig) +
    guides(colour = F) +
    facet_wrap(~trait_with_pig, ncol = 3) +
    theme_bw() +
    ggtitle("ECDF") +
    labs(colour = "Trait",
         x = expression(F[ST]),
         y = NULL)
```

```{r, fig.height=8, fig.width=12, fig.cap = "Unfaceted"}
fst_pegas_df_clumps %>% 
  # filter for clumped index SNPs
  dplyr::filter(FILTER == "CLUMPED") %>%

  # create new column combining skin pigmentation traits
  dplyr::mutate(trait_with_pig = dplyr::recode(trait,
                                               "skin pigmentation" = "skin pigmentation / measurement",
                                               "skin pigmentation measurement" = "skin pigmentation / measurement")) %>% 
  # remove all pigmentation traits other than skin pigmentation and skin pigmentation measurement
  dplyr::filter(trait_with_pig %in% traits_with_pig) %>% 
  # group by trait to take unique SNPs
  dplyr::group_by(trait_with_pig) %>% 
  # take unique IDs
  dplyr::distinct(ID, .keep_all = T) %>%
  # order traits
  dplyr::mutate(trait_with_pig = factor(trait_with_pig, levels = traits_with_pig)) %>%
  # plot
  ggplot() +
    stat_ecdf(aes(FST_PEGAS, colour = trait_with_pig)) +
    scale_colour_manual(values = pal_primary_with_pig) +
    theme_bw() +
    ggtitle("ECDF") +
    labs(colour = "Trait",
         x = expression(F[ST]),
         y = NULL)
```

# Kolmogorov-Smirnov tests

Comparing $F_{ST}$ distributions of each trait against each other.

```{r, warning=F, message=F}
ks_list = fst_pegas_df_clumps %>% 
  # filter for clumped index SNPs
  dplyr::filter(FILTER == "CLUMPED") %>%

  # create new column combining skin pigmentation traits
  dplyr::mutate(trait_with_pig = dplyr::recode(trait,
                                               "skin pigmentation" = "skin pigmentation / measurement",
                                               "skin pigmentation measurement" = "skin pigmentation / measurement")) %>% 
  # remove all pigmentation traits other than skin pigmentation and skin pigmentation measurement
  dplyr::filter(trait_with_pig %in% traits_with_pig) %>% 
  # group by trait to take unique SNPs
  dplyr::group_by(trait_with_pig) %>% 
  # take unique IDs
  dplyr::distinct(ID, .keep_all = T) %>%
  # order traits
  dplyr::mutate(trait_with_pig = factor(trait_with_pig, levels = traits_with_pig)) %>% 
  # split by trait
  split(., f = .$trait_with_pig)

# Run KS test
ks_out = purrr::map(ks_list, function(TRAIT_1){
  lapply(ks_list, function(TRAIT_2){
    out = ks.test(TRAIT_1$FST_PEGAS,
                  TRAIT_2$FST_PEGAS)
    df = data.frame(D_STATISTIC = out$statistic,
                    P_VALUE = out$p.value)
  }) %>% 
    dplyr::bind_rows(.id = "TRAIT_2")
}) %>% 
  dplyr::bind_rows(.id = "TRAIT_1") %>% 
  # order TRAIT variables
  dplyr::mutate(TRAIT_1 = factor(TRAIT_1, levels = traits_with_pig),
                TRAIT_2 = factor(TRAIT_2, levels = traits_with_pig))

```

## D-statistic

```{r, fig.height=9, fig.width=10}
ks_out %>% 
  # plot
  ggplot() +
    geom_tile(aes(TRAIT_1, TRAIT_2, fill = D_STATISTIC)) + 
    scale_fill_viridis_c() +
    theme_bw() + 
    coord_fixed() +
    labs(x = NULL, y = NULL) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## P-value

```{r, fig.height=9, fig.width=10}
ks_out %>% 
  # plot
  ggplot() +
    geom_tile(aes(TRAIT_1, TRAIT_2, fill = P_VALUE)) + 
    scale_fill_viridis_c(option = "inferno") +
    theme_bw() + 
    coord_fixed() +
    labs(x = NULL, y = NULL) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


