---
title: "Introgression in the MIKK panel"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

We want to determine the extent to which the MIKK panel shows evidence of introgression with other medaka populations, specifically the Northern and Korean medaka strains.

# Setup

## Create directory structure and clone repo

(Working directory on EBI cluster: `/hps/research1/birney/users/ian/mikk_paper` - the same as for the LD analysis) 

```{bash, eval = F}
# move to working directory
cd /your/working/directory
# clone git repository if it's not already there
git clone https://github.com/Ian-Brettell/mikk_genome.git
```

## Create conda evironment

```{bash, eval = F}
conda env create \
  -n mikk_env \
  -f mikk_genome/code/config/conda_env.yml
  
conda activate mikk_env
```

# Obtain multiple alignment data 

```{bash, eval = F}
mkdir emfs
# download
wget -P emfs/ ftp://ftp.ensembl.org/pub/release-100/emf/ensembl-compara/multiple_alignments/47_fish.epo/*

# unzip into new directory
mkdir emfs/unzipped

for i in $(find emfs/47_fish.epo.[0-9]*); do
  name=$(basename $i | cut -f3,4 -d'.');
  bsub "zcat $i > emfs/unzipped/$name";
done


# NOTE that file 17_1 is in a completely different format, with CIGAR strings instead of the normal SEQ, TREE, ID and DATA segments.
# want to remove all lines starting with $VAR, a space (make it two to be safe), and all consecutive blank lines (leaving single blank lines) 

# remove all lines starting with $VAR or two spaces, and all consecutive blank lines
sed '/\$VAR1/d; /\ \ /d' emfs/unzipped/17_1.emf | uniq -u > emfs/unzipped/tmp.txt
# replace old file
mv emfs/unzipped/tmp.txt emfs/unzipped/17_1.emf
```

## Tree plot

The tree file was copied and pasted from within `emfs/README.47`.

If generating the tree directly, the tips remove the spaces between Genus and species, so manually edit using regex to find spaces and replace them with "_":
`{bash} (?<=[a-z])( )(?=[a-z])`

```{r}
phylo_tree <- ape::read.tree(file = "~/Documents/Repositories/mikk_genome/data/20200921_47-fish-epo_tree.txt")
```

```{r, fig.width=11, fig.height=12.5}
# Colour all Oryzias
ids <- phylo_tree$tip.label[grep("Oryzias", phylo_tree$tip.label)]
# get indices of edges descending from MRCA (determined through trial and error)
oryzias_nodes <- seq(26, 33)
all_med_col <- ifelse(1:length(phylo_tree[["edge.length"]]) %in% oryzias_nodes, "#E84141", "black")
# set colours for tip labels
all_med_tip <- ifelse(phylo_tree$tip.label %in% ids, "#E84141", "black")
# plot
ape::plot.phylo(phylo_tree,
                use.edge.length = T,
                edge.color = all_med_col,
                tip.color = all_med_tip,
                font = 4)
```

```{r, eval = F}
# Save to repo
png(file= here("plots", "tree_all_oryzias.png"),
    width=22,
    height=25,
    units = "cm",
    res = 400)
ape::plot.phylo(phylo_tree,
                use.edge.length = T,
                edge.color = all_med_col,
                tip.color = all_med_tip,
                font = 4)
dev.off()
```

[Karyoplot Markdown App](20201125_karyoplot_app.Rmd)
